Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-07-02 talk/icooolps2009/talk/talk.tex
first round of simplification

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item mixed-mode execuction environment
    \item at first, everything is interpreted
    \item lightweight profiling to discover hot loops
    \item code generation only for common paths of hot loops
    \item when a hot loop is discovered, start to produce a trace
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Tracing}
    \begin{itemize}
    \item a \emph{trace} is a sequential list of operations
    \item a trace is produced by recording every operation the interpreter executes
    \item tracing ends when the tracer sees a position in the program it has seen before
    \item to identify these places, the \emph{position key} is used
    \item the position key encodes the current point of execution
    \item a trace thus corresponds to exactly one loop
    \item that means it ends with a jump to its beginning
    \end{itemize}
    \pause
    \begin{block}{Guards}
        \begin{itemize}
        \item the trace is only one of the possible code paths through the loop
        \item at places where the path \emph{could} diverge, a guard is placed
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Code Generation and Execution}
    \begin{itemize}
    \item being linear, the trace can easily be turned into machine code
    \item the machine code can be immediately executed
    \item execution stops when a guard fails
    \item after a guard failure, go back to interpreting program
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{(Dis-)Advantages of Tracing JITs}
    \begin{block}{Good Points of the Approach}
        \begin{itemize}
        \item easy and fast machine code generation: needs so support only one path
        \item interpreter does a lot of the work
        \item can be added to an existing interpreter unobtrusively
        \item automatic inlining
        \item produces very little code
        \end{itemize}
    \end{block}
    \pause
    \begin{block}{Bad Points of the Approach}
        \begin{itemize}
        \item unclear whether assumptions are true often enough
        \item switching between interpretation and machine code execution takes time
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (usually only) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: two iterations of the dispatch loop will usually take very different code paths
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter runs on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

@@

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE}}
\begin{verbatim}
loop_start(a0, regs0, bytecode0, pc0)
# MOV_R_A 0
opcode0 = strgetitem(bytecode0, pc0)
pc1 = int_add(pc0, Const(1))
guard_value(opcode0, Const(2))
n1 = strgetitem(bytecode0, pc1)
pc2 = int_add(pc1, Const(1))
a1 = call(Const(<* fn list_getitem>), regs0, n1)
# DECR_A
...
# MOV_A_R 0
...
# MOV_R_A 2
...
# ADD_R_TO_A 1
...
# MOV_A_R 2
...
# MOV_R_A 0
...
# JUMP_IF_A 4
opcode6 = strgetitem(bytecode0, pc13)
pc14 = int_add(pc13, Const(1))
guard_value(opcode6, Const(3))
target0 = strgetitem(bytecode0, pc14)
pc15 = int_add(pc14, Const(1))
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0, bytecode0, target0)
\end{verbatim}
}

@@

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE} With Constant-Folding}
\begin{verbatim}
loop_start(a0, regs0)
# MOV_R_A 0
a1 = call(Const(<* fn list_getitem>), regs0, Const(0))
# DECR_A
a2 = int_sub(a1, Const(1))
# MOV_A_R 0    
call(Const(<* fn list_setitem>), regs0, Const(0), a2)
# MOV_R_A 2
a3 = call(Const(<* fn list_getitem>), regs0, Const(2))
# ADD_R_TO_A  1
i0 = call(Const(<* fn list_getitem>), regs0, Const(1))
a4 = int_add(a3, i0)
# MOV_A_R 2
call(Const(<* fn list_setitem>), regs0, Const(2), a4)
# MOV_R_A 0
a5 = call(Const(<* fn list_getitem>), regs0, Const(0))
# JUMP_IF_A 4
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0)
\end{verbatim}
}

>>>>

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item mixed-mode execuction environment
    \item at first, everything is interpreted
    \item lightweight profiling to discover hot loops
    \item code generation only for common paths of hot loops
    \item when a hot loop is discovered, start to produce a trace
    \item when a full loop is traced, the trace is converted to machine code
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{(Dis-)Advantages of Tracing JITs}
    \begin{block}{Good Points of the Approach}
        \begin{itemize}
        \item easy and fast machine code generation: needs so support only one path
        \item (things are more complex, but let's ignore that for now)
        \item interpreter does a lot of the work
        \item can be added to an existing interpreter unobtrusively
        \item automatic inlining
        \item produces comparatively little code
        \end{itemize}
    \end{block}
    \pause
    \begin{block}{Bad Points of the Approach}
        \begin{itemize}
        \item unclear whether assumptions are true often enough
        \item switching between interpretation and machine code execution takes time
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: two iterations of the dispatch loop will usually take very different code paths
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter runs on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

@@

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE}}
\begin{verbatim}
loop_start(a0, regs0, bytecode0, pc0)
# MOV_R_A 0
opcode0 = strgetitem(bytecode0, pc0)
pc1 = int_add(pc0, Const(1))
guard_value(opcode0, Const(2))
n1 = strgetitem(bytecode0, pc1)
pc2 = int_add(pc1, Const(1))
a1 = list_getitem(regs0, n1)
# DECR_A
...
# MOV_A_R 0
...
# MOV_R_A 2
...
# ADD_R_TO_A 1
...
# MOV_A_R 2
...
# MOV_R_A 0
...
# JUMP_IF_A 4
opcode6 = strgetitem(bytecode0, pc13)
pc14 = int_add(pc13, Const(1))
guard_value(opcode6, Const(3))
target0 = strgetitem(bytecode0, pc14)
pc15 = int_add(pc14, Const(1))
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0, bytecode0, target0)
\end{verbatim}
}

@@

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE} With Constant-Folding}
\begin{verbatim}
loop_start(a0, regs0)
# MOV_R_A 0
a1 = list_getitem(regs0, Const(0))
# DECR_A
a2 = int_sub(a1, Const(1))
# MOV_A_R 0    
list_setitem(regs0, Const(0), a2)
# MOV_R_A 2
list_getitem(regs0, Const(2))
# ADD_R_TO_A  1
i0 = list_getitem(regs0, Const(1))
a4 = int_add(a3, i0)
# MOV_A_R 2
list_setitem(regs0, Const(2), a4)
# MOV_R_A 0
a5 = list_getitem(regs0, Const(0))
# JUMP_IF_A 4
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0)
\end{verbatim}
}
